---
title: "607_Window_Functions_EC"
author: 'jfor: Josh Forster'
date: "2022-09-12"
output: html_document
---
### Load Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RcppRoll)
```

## R Markdown


####Load Google and Tesla prices over past year (rolling 12 Month Window)

```{r }
goog <- read.csv('https://raw.githubusercontent.com/jforster19/DATA607/main/GOOG.csv',colClasses=c("Date",rep("numeric",6)))
tsla <- read.csv('https://raw.githubusercontent.com/jforster19/DATA607/main/TSLA.csv',colClasses=c("Date",rep("numeric",6)))
```


### Use window functions (in SQL or dplyr) to calculate the year-to-date average and the six-day moving averages for each item.
```{r pressure, echo=FALSE}

#found package for 6 day moving windows/based on tidyverse guidance https://datacornering.com/running-moving-rolling-average-in-r-dplyr/
goog$symbol <- 'GOOG'
tsla$symbol <- 'TSLA'
stocks <- bind_rows(goog,tsla)

stocks <- stocks %>%
    select(symbol, Date,Close) %>%
    group_by(symbol) %>%
    arrange(Date) %>%
    mutate(mov_avg=cummean(Close),
           six_day_movavg = roll_mean(Close,n = 6, align = "right", fill = NA)) %>%
    arrange(symbol,Date)
stocks           
```
The moving average calculation uses RCppRoll to get the rolling window of time that is needed, while the cumulative ytd average is calculated with dplyr.
